---
title: "Dynamic Bubble Plots"
author: "Kira Goff"
date: "2026-07-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r clear-workspace}
rm(list = ls(all = TRUE))
```



```{r}
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(ggtext)
library(paletteer)
```



```{r load-data}
# csv files as a list - can take one or more. should have standard taxonomy columns
# and then samples with whole numbers
file_paths <- c(
  "/mnt/work-drive/kira/nanopore16s/run1/final_clean_taxa_table.csv",
  "/mnt/work-drive/kira/nanopore16s/run2/final_clean_taxa_table.csv"
)

# read and merge dynamically 
data_list <- lapply(file_paths, read.csv)

# you might have to change these column names based on the database you used, look at your files
tax_cols <- c("OTU", # change this if your equivalent column is named anything else
              "Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species") 
common_cols <- Reduce(intersect, lapply(data_list, colnames))
join_cols <- intersect(tax_cols, common_cols)

if (length(data_list) == 1) {
  combined_data <- data_list[[1]]
} else {
  combined_data <- Reduce(function(x, y) {
    full_join(x, y, by = join_cols, relationship = "many-to-many")
  }, data_list)
}

# Fix NAs in counts/abundance columns
combined_data <- combined_data %>%
  mutate(across(where(is.numeric), ~replace_na(., 0)))

sample_cols <- colnames(combined_data)[sapply(combined_data, is.numeric)]
```

# fixy fixy

```{r taxa-filtering-and-collapse}
# --- BIOLOGICAL FILTERS & TAXONOMY OPTIONS ---
target_tax_level       <- "Genus"     # Phylum, Class, Order, Family, Genus
top_n_taxa             <- 25          # top n taxa
bubble_thresh          <- 0.001       # don't plot bubbles below x % relative abundance
filter_strategy        <- "sample"    # are the top n in all samples ("global") or for each sample ("sample")?
use_relative_abundance <- TRUE        # for absolute numbers, use FALSE
within_phylum_sort     <- "abundance" # "abundance", "alphabet", "occupancy" (sort by presence/absence)

# --- NEW SORTING OPTION ---
group_by_phylum        <- TRUE # TRUE (sort WITHIN phyla), FALSE (sort across all taxa)
# if group_by_phylum is FALSE, you must ALSO set facet_by_phylumn to FALSE

# --- UNCLASSIFIED TAXA CONTROLS ---
# Options: 
#   "parent_failed" -> e.g., "Bacilli Order" or "Lactobacillaceae Genus" 
#   "unclass_of"    -> e.g., "Unclassified Genus of Lactobacillaceae" 
#   "prefix"        -> e.g., "Uncl. Lactobacillaceae"
#   "suffix"        -> e.g., "Lactobacillaceae uncl."
#   "clean"         -> e.g., "Lactobacillaceae"
uncl_style             <- "parent_failed"      
fallback_label         <- "Unknown Bacteria" 

# --- OPTIONAL: METADATA SAMPLE SORTING ---
use_metadata_sort   <- FALSE              # TRUE to sort X-axis by metadata, FALSE to keep default
#metadata_file     <- "/mnt/work-drive/kira/rproj-rnaseq/results2/testmeta.csv"    # Path to your metadata file
#metadata_sample_col <- "Sample"        # Column in metadata matching your CSV column headers
#metadata_sort_col   <- "Order"       # Column in metadata you want to sort samples by (e.g., "Day", "Group")

# If not using, and you don't have a metadata file
# hash out metadata_file through metadata_sort_col
# so it stops freaking out when it fails to find a file you don't need
# -----

# --- INITIALIZE COLS TO PLOT ---
# We make a copy of sample_cols here. If metadata sorting is off, plot_cols remains untouched 
plot_cols <- sample_cols 

# Validate metadata selection immediately if active
if (use_metadata_sort) {
  if (!exists("metadata_file") || !file.exists(metadata_file)) {
    stop("Metadata file is either missing or you forgot to comment it back on")
  }
  if (!exists("metadata_sample_col") || !exists("metadata_sort_col")) {
    stop("You forgot to turn the metadata columns off.")
  }
  
  # Load metadata (auto-detects tab vs comma)
  meta_df <- read.table(metadata_file, header = TRUE, 
                        sep = ifelse(grepl("\\.tsv$\vert{}\\.txt$", tolower(metadata_file)), "\t", ","), 
                        stringsAsFactors = FALSE)
  
  if (!metadata_sample_col %in% colnames(meta_df)) {
    stop(metadata_sample_col, "' is not in the metadata file.")
  }
  if (!metadata_sort_col %in% colnames(meta_df)) {
    stop("Sort column '", metadata_sort_col, "' does not exist in the metadata file.")
  }
  
  # Extract sorted sample order from metadata
  ordered_samples <- meta_df %>%
    arrange(.data[[metadata_sort_col]]) %>%
    pull(.data[[metadata_sample_col]])
  
  # Keep only samples that actually exist in our raw data's sample_cols
  ordered_samples <- ordered_samples[ordered_samples %in% sample_cols]
  
  if (length(ordered_samples) == 0) {
    stop("Ooops, there are no samples names in your metadata that match your dataframe.")
  }
  
  # Append any dataset columns that were missing from metadata
  missing_samples <- setdiff(sample_cols, ordered_samples)
  if (length(missing_samples) > 0) {
    warning("Note: ", length(missing_samples), " sample(s) were missing from metadata. Appending to end.")
    ordered_samples <- c(ordered_samples, missing_samples)
  }
  
  # Save the finalized order
  plot_cols <- ordered_samples
}

# --- FIX UNCLASSIFIED TAXONOMY ---
rescue_taxonomy <- function(df, level, style = "parent_failed", fallback = "Unknown Bacteria") {
  levels_order <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
  target_idx <- which(levels_order == level)
  
  tax_matrix <- as.matrix(df[, levels_order[1:target_idx], drop = FALSE])
  clean_labels <- character(nrow(df))
  
  for (r in 1:nrow(df)) {
    current_val <- tax_matrix[r, target_idx]
    
    if (is.na(current_val) || current_val == "" || str_detect(tolower(current_val), "unclassified")) {
      
      # 1. Find the FIRST level that failed (moving left-to-right)
      first_unclass_idx <- NA
      for (c in 1:target_idx) {
        val <- tax_matrix[r, c]
        if (is.na(val) || val == "" || str_detect(tolower(val), "unclassified")) {
          first_unclass_idx <- c
          break
        }
      }
      
      # 2. Grab last known parent
      resolved_name <- fallback
      if (!is.na(first_unclass_idx) && first_unclass_idx > 1) {
        for (p_idx in (first_unclass_idx - 1):1) {
          parent_val <- tax_matrix[r, p_idx]
          if (!is.na(parent_val) && parent_val != "" && !str_detect(tolower(parent_val), "unclassified")) {
            
            failed_level_name <- levels_order[first_unclass_idx]
            
            resolved_name <- switch(style,
              "parent_failed" = paste(parent_val, failed_level_name),
              "unclass_of"    = paste("Unclassified", failed_level_name, "of", parent_val), 
              "suffix"        = paste(parent_val, "uncl."),
              "prefix"        = paste("Uncl.", parent_val),
              "clean"         = parent_val,
              paste(parent_val, "uncl.") 
            )
            break
          }
        }
      }
      clean_labels[r] <- resolved_name
    } else {
      if (level %in% c("Genus", "Species")) {
        clean_labels[r] <- paste0("*", current_val, "*")
      } else {
        clean_labels[r] <- current_val
      }
    }
  }
  df$Clean_Tax_Label <- clean_labels
  return(df)
}

processed_data <- rescue_taxonomy(
  combined_data, 
  level = target_tax_level, 
  style = uncl_style, 
  fallback = fallback_label
)

# Collapse duplicates (use plot_cols to keep columns safe)
collapsed_data <- processed_data %>%
  group_by(Phylum, Clean_Tax_Label) %>%
  summarise(across(all_of(plot_cols), sum), .groups = 'drop')

# convert to % if relative abundance toggled
if (use_relative_abundance) {
  collapsed_data <- collapsed_data %>%
    mutate(across(all_of(plot_cols), ~ (. / sum(.)) * 100))
  abundance_measure_text <- "Relative Abundance (%)"
  abundance_file_stamp   <- "relative"
} else {
  collapsed_data <- collapsed_data %>%
    mutate(across(all_of(plot_cols), ~ .))
  abundance_measure_text <- "Raw Read Count"
  abundance_file_stamp   <- "absolute"
}

# filter to top X taxa
if (filter_strategy == "global") {
  bubble_filtered <- collapsed_data %>%
    mutate(Total_Abundance = rowSums(across(all_of(plot_cols)))) %>%
    arrange(desc(Total_Abundance)) %>%
    slice_head(n = top_n_taxa) %>%
    pivot_longer(cols = all_of(plot_cols), names_to = "Sample", values_to = "Abundance")
} else {
  bubble_filtered <- collapsed_data %>%
    pivot_longer(cols = all_of(plot_cols), names_to = "Sample", values_to = "Abundance") %>%
    group_by(Sample) %>%
    slice_max(order_by = Abundance, n = top_n_taxa, with_ties = FALSE) %>%
    ungroup()
}

bubble_plot_data <- bubble_filtered %>% filter(Abundance > bubble_thresh)

# Handle phylum sorting
bubble_plot_data <- bubble_plot_data %>%
  group_by(Phylum) %>%
  mutate(Phylum_Total = sum(Abundance)) %>%
  group_by(Phylum, Clean_Tax_Label) %>%
  mutate(Taxon_Total = sum(Abundance),
         Taxon_Occupancy = sum(Abundance > 0)) %>%
  ungroup()

# Sort dataframe 
# Determine sort hierarchy based on whether we are grouping by phylum
if (group_by_phylum) {
  # Sorts within Phyla first 
  if (within_phylum_sort == "abundance") {
    bubble_plot_data <- bubble_plot_data %>% arrange(desc(Phylum_Total), desc(Taxon_Total))
  } else if (within_phylum_sort == "alphabet") {
    bubble_plot_data <- bubble_plot_data %>% arrange(Phylum, Clean_Tax_Label)
  } else if (within_phylum_sort == "occupancy") {
    bubble_plot_data <- bubble_plot_data %>% arrange(desc(Phylum_Total), desc(Taxon_Occupancy))
  }
} else {
  # Sorts globally by individual taxon abundance, ignoring phylum boundaries
  bubble_plot_data <- bubble_plot_data %>% arrange(desc(Taxon_Total))
}

# Extract factors in sorted order
ordered_phyla <- unique(bubble_plot_data$Phylum)
y_levels      <- rev(unique(bubble_plot_data$Clean_Tax_Label))

# Apply the factors safely
bubble_plot_data <- bubble_plot_data %>%
  mutate(
    Phylum = factor(Phylum, levels = ordered_phyla),
    Clean_Tax_Label = factor(Clean_Tax_Label, levels = y_levels)
  )

# Set the factor level order for the Sample column to match plot_cols order exactly
bubble_plot_data <- bubble_plot_data %>%
  mutate(Sample = factor(Sample, levels = plot_cols))
```

Some palettes that support a larger number of colours:
    pals::polychrome (Up to 36 colors): Built specifically dozens of categories
    pals::alphabet (Up to 26 colors): Highly distinct, contrasting colors
    pals::kelly (Up to 22 colors): Maximize contrast; excellent for making individual bubbles pop
    ggsci::ucscgb_pal (Up to 26 colors)
    ggsci::d3_category20 (20 colors)

But remember, we're colour coding by phyla, so you don't need as many as you might think
Paletteer gallery: https://pmassicotte.github.io/paletteer_gallery/#discrete-palettes
Palette finder: https://r-graph-gallery.com/color-palette-finder.html 

```{r data-visualization-and-save}
# ----- VISUALIZATION OPTIONS -----
phylum_palette         <- "ggthemes::Tableau_20" # palette for paletteer 
plot_width             <- 11            
plot_height            <- 15            

facet_by_phylum        <- TRUE          # must be FALSE if group_by_phylum is FALSE
show_abundance_labels  <- FALSE          # do you want the % relative abundance printed on each bubble
legend_position        <- "right"        # "none", "bottom", "right"
bubble_max             <- 400           # adjust up for generally larger bubbles, down if overlapping

# ----------------------------------

# --- DYNAMIC BUBBLE SIZING ---
num_taxa    <- length(unique(bubble_plot_data$Clean_Tax_Label))
num_samples <- length(unique(bubble_plot_data$Sample))

# Shrink maximum bubble size as the grid gets more crowded
bubble_min_size <- max(1.5, min(4, 30 / num_taxa))
bubble_max_size <- max(4, min(18, bubble_max / num_taxa))

palette_clean_name <- str_replace(phylum_palette, "::", "_")

# --- THE PLOT THICKENS ---

# ggplot base
p <- ggplot(bubble_plot_data, aes(x = Sample, y = Clean_Tax_Label, size = Abundance, color = Phylum)) +
  geom_point(alpha = 0.85) +
  scale_size_continuous(range = c(bubble_min_size, bubble_max_size), name = abundance_measure_text) +
  scale_color_paletteer_d(phylum_palette, name = "Phylum")

if (show_abundance_labels) {
  p <- p + geom_text(
    aes(label = if(use_relative_abundance) paste0(round(Abundance, 1), "%") else round(Abundance, 0)),
    color = "black",
    size = 3.0,
    fontface = "bold",
    show.legend = FALSE
  )
}

if (facet_by_phylum) {
  p <- p + 
    theme_bw() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 11, face = "bold"),
      # Target both general and specific left y-axis elements
      axis.text.y = ggtext::element_markdown(size = 10),
      axis.text.y.left = ggtext::element_markdown(size = 10),
      strip.text.y = element_text(angle = 0, face = "bold", size = 9),
      strip.background = element_rect(fill = "gray95"),
      panel.grid.major = element_line(color = "gray95", linewidth = 0.5)
    ) +
    labs(x = "Samples", y = "") +
    facet_grid(Phylum ~ ., scales = "free_y", space = "free_y")
  
  layout_type <- "boxed"
} else {
  p <- p + 
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 11, face = "bold"),
      # Target both general and specific left y-axis elements to force-override theme_minimal
      axis.text.y = ggtext::element_markdown(size = 10), 
      axis.text.y.left = ggtext::element_markdown(size = 10), 
      axis.title = element_text(size = 12, face = "bold"),
      panel.grid.major.y = element_blank(),
      panel.grid.minor.y = element_blank(),
      panel.grid.major.x = element_line(color = "gray95", linewidth = 0.5),
      panel.border = element_rect(color = "gray80", fill = NA, linewidth = 1)
    ) +
    labs(
      x = "Samples",
      y = paste("Bacterial", target_tax_level),
      title = paste(target_tax_level, "Level Profile"),
      subtitle = paste("Sorted by:", within_phylum_sort)
    )
  
  layout_type <- "flat"
}

# Legend settings
if (legend_position == "none") {
  p <- p + theme(legend.position = "none")
} else if (legend_position == "bottom") {
  p <- p + 
    guides(
      color = "none", 
      size = guide_legend(direction = "horizontal", title.position = "top", title.hjust = 0.5)
    ) +
    theme(
      legend.position = "bottom",
      legend.background = element_blank(),
      legend.key = element_blank()
    )
} else if (legend_position == "right") {
  p <- p + theme(
    legend.position = "right",
    legend.title = element_text(face = "bold")
  )
}

print(p)

# --- DYNAMIC EXPORT ---

# Capture date in YYYYMMDD format
date_stamp <- format(Sys.Date(), "%Y%m%d")

# replace variables with something that will make sense in the future
layout_label  <- if (facet_by_phylum) "boxed" else "flat"
label_suffix  <- if (show_abundance_labels) "_withLabels" else ""
tax_sorting_label <- if (group_by_phylum) "byPhylum" else "globalSort" # y sort

# append "_sortBy[Column]" if metadata sorting is used
sample_sorting_suffix <- if (use_metadata_sort && exists("metadata_sort_col")) {
  paste0("_sortBy", metadata_sort_col)
} else {
  ""
}

# Assemble file name
output_filename <- paste0(
  "bubbleplot_",
  layout_label, "_",
  target_tax_level, "_",
  tax_sorting_label, 
  sample_sorting_suffix, "_",  # Appends metadata sort info right after tax sort
  filter_strategy, "_",
  "top", top_n_taxa, "_",
  abundance_file_stamp,
  label_suffix, "_",
  date_stamp, ".png"
)

# Save plot
ggsave(
  filename = output_filename, 
  plot     = p, 
  width    = plot_width, 
  height   = plot_height, 
  dpi      = 300
)

# Success message
message("Huzzah! Plot saved as: ", output_filename)
```